---
title: "R Notebook"
output: html_notebook
---
Load standard packages
```{r, message=FALSE, warning=FALSE, include=FALSE}
rm(list = ls())
library(tidyverse)
library(scales)
library(broom)
library(tidyheatmaps)
library(clusterProfiler)
library(org.Hs.eg.db)
library(scales)
library(ggdendro)
library(ggrepel)
library(lavaan)
library(enrichplot)
library(lme4)
library(lmerTest)
library(RColorBrewer)
library(broom.mixed)
setwd("~/Desktop/R_work/Standards")
list.files() %>% map(source)
```
Define directories
```{r}
dir_raw <- "path/to/your/directory"
dir_plots <- "path/to/your/directory"
dir_stats <- "path/to/your/directory"
dir_obj <- "path/to/your/directory"
```
load data
```{r}
dat <- read_csv("/path/to/your/file.csv")
```
define parameters
```{r}
imaging_cor1 <- c(paste("mk_adni_suvr_infcg_braak", seq(1:6), sep = ""), "mk_adni_suvr_infcg_metaroi", "braak_stage", "nav_adni_suvr_fullcg_neocortex", "hp_vol")
cogn_cor1 <- c("cdr", "cdr_sob", "mmse")
fluid_cor <- c("csf_ab40_pg_ml__lumip1", "csf_ab42_pg_ml__lumip1", "csf_ttau_pg_ml__lumip1", "csf_ptau181_pg_ml__simoa1", "csf_ptau217_pg_ml__simoa", "csf_ptau231_pg_ml__simoa1", "csf_ptau235_pg_ml__simoa", "csf_ptau202_pg_ml__simoa2", "csf_ptau205_pg_ml__simoa", "csf_taun224_pg_ml_", "csf_tau368_pg_ml_", "csf_nta_pg_ml__simoa", "csf_nfl_pg_ml_")
```
analyze seperately for diagnosis
```{r, fig.width=10, fig.height=10}
dat_include <- 
  dat %>% 
  filter(tissue == "csf") %>% 
  dplyr::select(data_f1) %>% 
  unnest %>% 
  filter(!is.na(diagnosis_v1), protein != "SampleQC") %>% 
  mutate(
    diagnosis_v2 = if_else(diagnosis_v1 == "Other" & mmse >= 27, "CN", diagnosis_v1),
    comps1 = if_else(diagnosis_v2 == "AD", "yes", NA), #AD vs CU
    comps1 = if_else(diagnosis_v2 %in% c("CNY"), "no", comps1),
    comps2 = if_else(diagnosis_v2 == "AD", "yes", NA), #AD vs MCI
    comps2 = if_else(diagnosis_v2 %in% c("MCI"), "no", comps2),
    comps3 = if_else(diagnosis_v2 == "MCI", "yes", NA), #MCI vs CU
    comps3 = if_else(diagnosis_v2 %in% c("CNY"), "no", comps3),
    comps4 = if_else(diagnosis_v2 == "CN", "yes", NA), #MCI vs CU
    comps4 = if_else(diagnosis_v2 %in% c("CNY"), "no", comps2),
    comps5 = if_else(diagnosis_v2 == "AD", "yes", "no"), #AD vs non AD,
    comps6 = if_else(diagnosis_v2 == "CN" & Amy == "A+", "yes", NA), #CU A+ vs CU A-
    comps6 = if_else(diagnosis_v2 == "CN" & Amy == "A-", "no", comps6), #CU A+ vs CU A-
    comps7 = if_else(diagnosis_v2 == "AD", "yes", NA), #AD vs CU
    comps7 = if_else(diagnosis_v2 %in% c("CN"), "no", comps7),
    comps8 = if_else(diagnosis_v2 == "MCI", "yes", NA), #MCI vs CU
    comps8 = if_else(diagnosis_v2 %in% c("CN"), "no", comps8),
    comps9 = if_else(diagnosis_v2 %in% c("MCI", "AD") & Amy == "A+", "yes", NA), #CI A+ vs CU A-
    comps9 = if_else(diagnosis_v2 %in% c("MCI", "AD") & Amy == "A-", "no", comps9), #CI A+ vs CU A-
    comps10 = if_else(diagnosis_v2 %in% c("MCI", "AD") & Amy == "A+", "yes", NA), #A+ vs A- with CUY excluded
    comps10 = if_else(diagnosis_v2 %in% c("CN") & Amy == "A-", "no", comps10), #A+ vs A- with CUY excluded
    comps11 = if_else(diagnosis_v2 %in% c("CN") & Amy == "A-", "yes", NA), #CN A- vs CNY
    comps11 = if_else(diagnosis_v2 %in% c("CNY"), "no", comps11), #CN A- vs CNY
    comps12 = if_else(diagnosis_v2 %in% c("MCI", "AD") & Amy == "A+", "yes", NA),
    comps12 = if_else(diagnosis_v2 %in% c("CN") & Amy == "A+", "no", comps12)
    )
# make t test -----
dat_include_test <- 
  dat_include %>% 
  group_by(protein) %>% 
  nest() %>% 
  mutate(
    #comp1: AD vs CUY
    comps1_p = map(data, function(x) x %>% filter(!is.na(comps1)) %>% compare_means(data = ., formula = Value_lod~comps1, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps1_padj = p.adjust(comps1_p, n = 122, method = "fdr"),
    comps1_log2fc = map(data, function(x) mean((x %>% filter(comps1 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps1 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp2: AD vs MCI
    comps2_p = map(data, function(x) x %>% filter(!is.na(comps2)) %>% compare_means(data = ., formula = Value_lod~comps2, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps2_padj = p.adjust(comps2_p, n = 122, method = "fdr"),
    comps2_log2fc = map(data, function(x) mean((x %>% filter(comps2 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps2 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp3: MCI vs CUY
    comps3_p = map(data, function(x) x %>% filter(!is.na(comps3)) %>% compare_means(data = ., formula = Value_lod~comps3, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps3_padj = p.adjust(comps3_p, n = 122, method = "fdr"),
    comps3_log2fc = map(data, function(x) mean((x %>% filter(comps3 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps3 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp4: CU vs CUY
    comps4_p = map(data, function(x) x %>% filter(!is.na(comps4)) %>% compare_means(data = ., formula = Value_lod~comps4, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps4_padj = p.adjust(comps4_p, n = 122, method = "fdr"),
    comps4_log2fc = map(data, function(x) mean((x %>% filter(comps4 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps4 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp5: AD vs non AD
    comps5_p = map(data, function(x) x %>% filter(!is.na(comps5)) %>% compare_means(data = ., formula = Value_lod~comps5, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps5_padj = p.adjust(comps5_p, n = 122, method = "fdr"),
    comps5_log2fc = map(data, function(x) mean((x %>% filter(comps5 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps5 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp6: CU A+ vs CU A-
    comps6_p = map(data, function(x) x %>% filter(!is.na(comps6)) %>% compare_means(data = ., formula = Value_lod~comps6, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps6_padj = p.adjust(comps6_p, n = 122, method = "fdr"),
    comps6_log2fc = map(data, function(x) mean((x %>% filter(comps6 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps6 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp7: AD vs CUY
    comps7_p = map(data, function(x) x %>% filter(!is.na(comps7)) %>% compare_means(data = ., formula = Value_lod~comps7, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps7_padj = p.adjust(comps7_p, n = 122, method = "fdr"),
    comps7_log2fc = map(data, function(x) mean((x %>% filter(comps7 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps7 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp8: MCI vs CUY
    comps8_p = map(data, function(x) x %>% filter(!is.na(comps8)) %>% compare_means(data = ., formula = Value_lod~comps8, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps8_padj = p.adjust(comps8_p, n = 122, method = "fdr"),
    comps8_log2fc = map(data, function(x) mean((x %>% filter(comps8 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps8 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp9: CI A+ vs CI A-
    comps9_p = map(data, function(x) x %>% filter(!is.na(comps9)) %>% compare_means(data = ., formula = Value_lod~comps9, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps9_padj = p.adjust(comps9_p, n = 122, method = "fdr"),
    comps9_log2fc = map(data, function(x) mean((x %>% filter(comps9 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps9 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp10: CI A+ vs CU A-
    comps10_p = map(data, function(x) x %>% filter(!is.na(comps10)) %>% compare_means(data = ., formula = Value_lod~comps10, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps10_padj = p.adjust(comps10_p, n = 122, method = "fdr"),
    comps10_log2fc = map(data, function(x) mean((x %>% filter(comps10 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps10 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp11: CU A- vs CUY
    comps11_p = map(data, function(x) x %>% filter(!is.na(comps11)) %>% compare_means(data = ., formula = Value_lod~comps11, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps11_padj = p.adjust(comps11_p, n = 122, method = "fdr"),
    comps11_log2fc = map(data, function(x) mean((x %>% filter(comps11 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps11 == "no") %>% pull(Value_lod)))) %>% unlist(),
    #comp12: CU A+ vs CI A+
    comps12_p = map(data, function(x) x %>% filter(!is.na(comps12)) %>% compare_means(data = ., formula = Value_lod~comps12, method = "t.test", p.adjust.method = "fdr") %>% pull(p.adj)) %>% unlist(),
    comps12_padj = p.adjust(comps12_p, n = 122, method = "fdr"),
    comps12_log2fc = map(data, function(x) mean((x %>% filter(comps12 == "yes") %>% pull(Value_lod))) - mean((x %>% filter(comps12 == "no") %>% pull(Value_lod)))) %>% unlist(),
    )

# make volcano plots -----
#make volcano plots
plot_volcano <- 
  dat_include_test %>% 
  dplyr::select(is.numeric) %>% 
  pivot_longer(-protein) %>% 
  separate(name, into = c("comp", "stat"), sep = "_") %>% 
  pivot_wider(id_cols = c(protein, comp), names_from = stat, values_from = value) %>% 
  mutate(
    nlogpadj = -log10(padj),
    color = if_else(log2fc < 0 & padj < 0.05, muted("blue"), "grey80"),
    color = if_else(log2fc > 0 & padj < 0.05, muted("red"), color),
    ) %>% 
  group_by(comp) %>% 
  arrange(padj) %>% 
  mutate(
    label = if_else(row_number() <= 15, protein, NA),
    label = if_else(padj < 0.05, label, NA)
    ) %>% 
  nest() %>% 
  mutate(
    volcano = pmap(list(data, comp), function(x,y){
      x %>% 
        ggplot(aes(x = log2fc, y = nlogpadj)) +
        geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
        geom_vline(xintercept = 0, linetype = "dashed") +
        geom_point(aes(fill = color), shape = 21, color = "white") +
        geom_text_repel(aes(label = label), size = 2.8, max.overlaps = Inf) +
        scale_x_continuous(limits = c(NA, NA)) +
        scale_fill_identity() +
        labs(title = y) +
        pl_msw0_nolegend_straight +
        plot_layout(widths = unit(4, "cm"), heights = unit(4, "cm"))
      })
    )
wrap_plots(plot_volcano$volcano)
ggsave(file.path(dir_plots, "volcano_plots_disease.pdf"), useDingbats = F)
#make GO term analysis -----
dat_go <- 
  plot_volcano %>% 
  filter(comp %in% c("comps1", "comps3", "comps4", "comps10", "comps11", "comps6")) %>% 
  dplyr::select(data) %>% 
  mutate(
    genes_signif = map(data, function(x) x %>% filter(padj < 0.05) %>% pull(protein)),
    go_bp = map(genes_signif, function(x) enrichGO(x, keyType = "SYMBOL", ont = "BP", pAdjustMethod = "fdr", pvalueCutoff = 1, "org.Hs.eg.db")),
    go_bp_res = map(go_bp, function(x) x@result %>% mutate(nlogpadj = -log10(p.adjust))),
    go_cc = map(genes_signif, function(x) enrichGO(x, keyType = "SYMBOL", ont = "CC", pAdjustMethod = "fdr", pvalueCutoff = 1, "org.Hs.eg.db")),
    go_cc_res = map(go_cc, function(x) x@result %>% mutate(nlogpadj = -log10(p.adjust)))
    )

# save volcano plots
plot_volcano %>% filter(comp== "comps10") %>% pull(data) %>% .[[1]] %>% dplyr::select(protein, padj, log2fc) %>% write.xlsx(file.path(dir_stats, "zz_f1a.xlsx"), overwrite = T)
plot_volcano %>% filter(comp== "comps6") %>% pull(data) %>% .[[1]] %>% dplyr::select(protein, padj, log2fc) %>% write.xlsx(file.path(dir_stats, "zz_f1i.xlsx"), overwrite = T)
plot_volcano %>% filter(comp== "comps4") %>% pull(data) %>% .[[1]] %>% dplyr::select(protein, padj, log2fc) %>% write.xlsx(file.path(dir_stats, "zz_af1a.xlsx"), overwrite = T)
plot_volcano %>% filter(comp== "comps3") %>% pull(data) %>% .[[1]] %>% dplyr::select(protein, padj, log2fc) %>% write.xlsx(file.path(dir_stats, "zz_af1b.xlsx"), overwrite = T)
plot_volcano %>% filter(comp== "comps1") %>% pull(data) %>% .[[1]] %>% dplyr::select(protein, padj, log2fc) %>% write.xlsx(file.path(dir_stats, "zz_af1c.xlsx"), overwrite = T)
plot_volcano %>% filter(comp== "comps11") %>% pull(data) %>% .[[1]] %>% dplyr::select(protein, padj, log2fc) %>% write.xlsx(file.path(dir_stats, "zz_af1f.xlsx"), overwrite = T)
```
make plots for pooled go terms (old version)
```{r}
#make plots for specific pooled go terms -----
dat_gol_line_v1 <- 
  dat_go %>% 
  dplyr::select(go_bp_res) %>% 
  unnest() %>% 
  filter(comp %in% c("comps4", "comps3", "comps1")) %>%  
  mutate(comp = factor(comp, levels = c("comps4", "comps3", "comps1"))) %>% 
  #make summary column
  mutate(
    bp = if_else(str_detect(Description, "migration"), "migration", NA),
    bp = if_else(str_detect(Description, "chemotaxis"), "chemotaxis", bp),
    bp = if_else(str_detect(Description, "ERK"), "ERK_signaling", bp),
    bp = if_else(str_detect(Description, "synaptic"), "synaptic_signaling", bp),
    bp = if_else(str_detect(Description, "calcium"), "calcium_signaling", bp),
    bp = if_else(str_detect(Description, "glia"), "glia", bp),
    bp = if_else(str_detect(Description, "cytokine"), "zcytokine", bp)
    ) %>% 
  filter(!is.na(bp)) %>% 
  separate(GeneRatio, into = c("g1", "g2"), sep = "/", remove = F) %>% 
  separate(BgRatio, into = c("b1", "b2"), sep = "/", remove = F) %>% 
  type_convert() %>% 
  mutate(effsize = ((g1/g2) / (b1/b2)),
         effsize2 = g1/g2,
         composite = FUN.scale_01(effsize) + FUN.scale_01(nlogpadj)) %>% 
  ungroup() %>% 
  add_row(comp = rep("CU", 7), bp = c("migration", "chemotaxis", "synaptic_signaling", "calcium_signaling", "ERK_signaling", "glia", "zcytokine"), nlogpadj = rep(0, 7)) %>% 
  add_row(comp = rep("CU", 7), bp = c("migration", "chemotaxis", "synaptic_signaling", "calcium_signaling", "ERK_signaling", "glia", "zcytokine"), effsize = rep(1, 7)) %>% 
  add_row(comp = rep("CU", 7), bp = c("migration", "chemotaxis", "synaptic_signaling", "calcium_signaling", "ERK_signaling", "glia", "zcytokine"), composite = rep(0, 7)) %>% 
  mutate(comp = factor(comp, levels = c("CU", "comps4", "comps3", "comps1")),
         comp_num = as.numeric(comp)
         )

#make plots for specific cytokines -----
dat_gol_line_v2 <- 
  dat_go %>% 
  dplyr::select(go_bp_res) %>% 
  unnest() %>% 
  filter(comp %in% c("comps4", "comps3", "comps1"),
         Description %in% c("response to type II interferon", "response to tumor necrosis factor", "response to interleukin-1")) %>%  
  separate(GeneRatio, into = c("g1", "g2"), sep = "/", remove = F) %>% 
  separate(BgRatio, into = c("b1", "b2"), sep = "/", remove = F) %>% 
  type_convert() %>% 
  mutate(effsize = ((g1/g2) / (b1/b2)),
         effsize2 = g1/g2,
         composite = FUN.scale_01(effsize) + FUN.scale_01(nlogpadj)) %>% 
  ungroup() %>% 
  add_row(comp = rep("CU", 3), Description = c("response to type II interferon", "response to tumor necrosis factor", "response to interleukin-1"), nlogpadj = rep(0, 3)) %>% 
  add_row(comp = rep("CU", 3), Description = c("response to type II interferon", "response to tumor necrosis factor", "response to interleukin-1"), effsize = rep(1, 3)) %>% 
  add_row(comp = rep("CU", 3), Description = c("response to type II interferon", "response to tumor necrosis factor", "response to interleukin-1"), composite = rep(0, 3)) %>% 
  mutate(comp = factor(comp, levels = c("CU", "comps4", "comps3", "comps1")),
         comp_num = as.numeric(comp)
         )

#make go term plots for individual comparisons -----
barplot(dat_go$go_bp[[1]], showCategory = 20) +
  plot_layout(widths = unit(5, "cm"), heights = unit(10, "cm"))
ggsave(file.path(dir_plots, "GO_barplot_ADvsCN.pdf"), useDingbats = F, width = unit(10, "cm"), height = unit(15, "cm"))
barplot(dat_go$go_bp[[2]], showCategory = 20) +
  plot_layout(widths = unit(5, "cm"), heights = unit(10, "cm"))
ggsave(file.path(dir_plots, "GO_barplot_MCIvsCN.pdf"), useDingbats = F, width = unit(10, "cm"), height = unit(15, "cm"))
barplot(dat_go$go_bp[[4]], showCategory = 20) +
  plot_layout(widths = unit(5, "cm"), heights = unit(10, "cm"))
ggsave(file.path(dir_plots, "GO_barplot_CNvsCNY.pdf"), useDingbats = F, width = unit(10, "cm"), height = unit(15, "cm"))
#save plots for CU Ab+ vs AB-
barplot(dat_go$go_bp[[3]], showCategory = c("synapse organization", "axonogenesis", "amyloid fibril formation", "NADH metabolic process", "axonal transport of mitochondrion", "canonical glycolysis")) +
  plot_layout(widths = unit(2, "cm"), heights = unit(4, "cm"))
ggsave(file.path(dir_plots, "go_barplot_CUAB+vsCUAB-.pdf"), useDingbats = F)

## Plot GO terms
dat_gol_line_v1 %>% 
  filter(comp != "CU") %>% 
  ggplot(aes(x = bp, y = nlogpadj, fill = bp)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  facet_wrap(~comp_num, ncol = 4) +
  pl_msw0_legend +
  scale_fill_manual(values = cbPalette[1:7]) +
  plot_layout(widths = unit(5, "cm"), heights = unit(2.5, "cm"))
ggsave(file.path(dir_plots, "New_GO_term_pooled_disease_significance.pdf"), useDingbats = F)

## Plot cytokines
dat_gol_line_v2 %>% 
  filter(comp != "CU") %>% 
  ggplot(aes(x = Description, y = nlogpadj, fill = Description)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  facet_wrap(~comp_num, ncol = 4) +
  pl_msw0_legend +
  scale_fill_manual(values = c(cbPalette[8:9], muted("red"))) +
  plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
ggsave(file.path(dir_plots, "NEW_GO_term_cytokine_disease_significance.pdf"), useDingbats = F)
```
compare CU AB- with CU AB+ by using AUCell
```{r}
#retrieve genes from go term analysis
go_interest <- 
  dat_go %>% 
  dplyr::select(go_bp_res) %>% 
  unnest() %>% 
  ungroup() %>% 
  filter(Description %in% c(c("chemotaxis", "leukocyte migration", "ERK1 and ERK2 cascade", "response to chemokine", "response to interleukin-1", "response to tumor necrosis factor", "response to type II interferon", "regulation of calcium-mediated signaling", "glial cell activation"))) %>% 
  group_by(Description, ID) %>% 
  nest() %>% 
  mutate(genes = map(data, function(x) x %>% pull(geneID) %>% paste(collapse = "/") %>% str_split("/") %>% unlist %>% unique))

#retrieve genes with go database
library(biomaRt)
biomart_genes <- 
  getBM(attributes = c('ensembl_gene_id', 'external_gene_name', 'go_id'), filters = 'go', values = go_interest$ID, mart = ensembl) %>% 
  filter(go_id %in% go_interest$ID) %>% 
  group_by(go_id) %>% 
  nest() %>% 
  mutate(gene = map(data, function(x) x$external_gene_name))
#merge go terms
go_interest <- 
  AnnotationDbi::select(org.Hs.eg.db,
                               keys = go_interest$ID,
                               columns = c("SYMBOL", "GENENAME"),
                               keytype = "GO") %>% 
  group_by(GO) %>% 
  nest() %>% 
  mutate(genes2 = map(data, function(x) x$SYMBOL)) %>% 
  right_join(go_interest, by = c("GO" = "ID")) %>% 
  mutate(genes3 = pmap(list(genes, genes2), function(x,y) c(x,y) %>% unique))
#make aucell analysis
aucell_mtx <- 
  dat %>% 
  filter(tissue == "csf") %>% 
  dplyr::select(data_f1) %>% 
  unnest %>% 
  dplyr::select(protein, id, Val_z) %>% 
  pivot_wider(id_cols = protein, names_from = id, values_from = Val_z) %>% 
  column_to_rownames("protein") %>% 
  as.matrix()

aucell_mtx_ranking <- AUCell_buildRankings(aucell_mtx, plotStats = F)
aucell_geneset <- go_interest$genes %>% set_names(go_interest$Description)
cells_AUC <- AUCell_calcAUC(aucell_geneset, aucell_mtx_ranking, aucMaxRank = 1000, normAUC = F)

#make aucell plot
plot_aucell <- 
  getAUC(cells_AUC) %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("id") %>% 
  pivot_longer(-id) %>% 
  type_convert() %>% 
  left_join(dat$data[[1]], by = "id") %>% 
  dplyr::select(name, diagnosis_v1, Amy, value) %>% 
  filter(!is.na(Amy), diagnosis_v1 != "Other") %>% 
  mutate(
    diagnosis_v3 = if_else(diagnosis_v1 %in% c("CN", "CNY"), "CN", "CI"),
    diagnosis_v4 = paste(diagnosis_v3, Amy, sep = "_"),
    diagnosis_v4 = factor(diagnosis_v4, levels = c("CN_A-", "CN_A+", "CI_A-", "CI_A+")),
    ) %>% 
  group_by(name) %>% 
  nest() %>% 
  mutate(
    plots_1 = pmap(list(data, name), function(x,y){
      x %>% 
        filter(diagnosis_v3 %in% "CN") %>% 
        mutate(value = FUN.scale_z(value)) %>% 
        ggplot(aes(x = diagnosis_v4, y = value, fill = diagnosis_v4)) +
        geom_violin(draw_quantiles = 0.5) +
        labs(title = y) +
        scale_fill_manual(values = cbPalette[c(1,2)]) +
        pl_msw0_legend +
        stat_compare_means(method = "t.test", comparisons = list(c("CN_A-", "CN_A+")), hide.ns = F, label = "p.signif") +
        plot_layout(widths = unit(1.25, "cm"), heights = unit(2.5, "cm"))
      }),
    plots_2 = pmap(list(data, name), function(x,y){
      x %>% 
        filter(diagnosis_v4 %in% c("CN_A-", "CI_A+")) %>%
        mutate(value = FUN.scale_z(value)) %>% 
        ggplot(aes(x = diagnosis_v4, y = value, fill = diagnosis_v4)) +
        geom_violin(draw_quantiles = 0.5) +
        labs(title = y) +
        scale_fill_manual(values = cbPalette[c(1,2)]) +
        pl_msw0_legend +
        stat_compare_means(method = "t.test", comparisons = list(c("CN_A-", "CI_A+")), hide.ns = F, label = "p.signif") +
        plot_layout(widths = unit(1.25, "cm"), heights = unit(2.5, "cm"))
      }),
    plots_3 = pmap(list(data, name), function(x,y){
      x %>% 
        filter(diagnosis_v4 %in% c("CN_A+", "CI_A+")) %>% 
        mutate(value = FUN.scale_z(value)) %>% 
        ggplot(aes(x = diagnosis_v4, y = value, fill = diagnosis_v4)) +
        geom_violin(draw_quantiles = 0.5) +
        labs(title = y) +
        scale_fill_manual(values = cbPalette[c(1,2)]) +
        pl_msw0_legend +
        stat_compare_means(method = "t.test", comparisons = list(c("CN_A+", "CI_A+")), hide.ns = F, label = "p.signif") +
        plot_layout(widths = unit(1.25, "cm"), heights = unit(2.5, "cm"))
      }),
    stats1 = map(data, function(x){
      x %>% 
        filter(diagnosis_v3 %in% "CN") %>% 
        mutate(value = FUN.scale_z(value)) %>% 
        compare_means(data = ., method = "t.test", formula =value~diagnosis_v4)}),
    stats2 = map(data, function(x){
      x %>% 
        filter(diagnosis_v4 %in% c("CN_A-", "CI_A+")) %>% 
        mutate(value = FUN.scale_z(value)) %>% 
        compare_means(data = ., method = "t.test", formula =value~diagnosis_v4)})
    )
```
check for gfap expression
```{r}
dat_include_test %>% 
  filter(protein == "GFAP") %>% 
  pull(data) %>% 
  .[[1]] %>% 
  filter(!is.na(comps10)) %>%
  ggplot(aes(x = comps10, y = Value_lod, fill = comps10)) +
  geom_violin(draw_quantiles = 0.5) +
  scale_fill_manual(values = cbPalette[c(1,2)]) +
  pl_msw0_legend +
  stat_compare_means(method = "t.test", comparisons = list(c("yes", "no")), hide.ns = F, label = "p.signif") +
  plot_layout(widths = unit(1.25, "cm"), heights = unit(2.5, "cm"))
ggsave(file.path(dir_plots, "gfap_CIA+vsCUA-.pdf"), useDingbats = F)
dat_include_test %>% 
  filter(protein == "GFAP") %>% 
  pull(data) %>% 
  .[[1]] %>% 
  filter(!is.na(comps6)) %>% 
  ggplot(aes(x = comps6, y = Value_lod, fill = comps6)) +
  geom_violin(draw_quantiles = 0.5) +
  scale_fill_manual(values = cbPalette[c(1,2)]) +
  pl_msw0_legend +
  stat_compare_means(method = "t.test", comparisons = list(c("yes", "no")), hide.ns = F, label = "p.signif") +
  plot_layout(widths = unit(1.25, "cm"), heights = unit(2.5, "cm"))
ggsave(file.path(dir_plots, "gfap_CUA+vsCUA-.pdf"), useDingbats = F)
```
calculate correlations and show them as heatmap
```{r, fig.height=10}
dat_prep <- 
  dat %>% 
  filter(tissue == "csf") %>% 
  dplyr::select(data_f1) %>% 
  unnest %>% 
  type_convert() %>% 
  ungroup() %>% 
  #filter(ATN %in% c("A-T-N-", "A+T-N-", "A+T+N-", "A+T+N+")) %>% 
  pivot_wider(id_cols = c("id", imaging_cor1, fluid_cor, cogn_cor1, "Gender", "age_at_mri", "APOE_num"), names_from = protein, values_from = Value_lod) %>% 
  filter(!is.na(nav_adni_suvr_fullcg_neocortex)) %>% 
  pivot_longer(ACHE:YWHAZ, names_to = "Protein", values_to = "Value_Protein") %>% 
  pivot_longer(cols = c(imaging_cor1, fluid_cor, cogn_cor1), names_to = "Imaging", values_to = "Value_Imaging") %>% 
  group_by(Protein, Imaging) %>% 
  nest()

#make correlation analysis  
cor_v1 <- 
  dat_prep %>% 
  mutate(
    cor_spearman = map(data, function(x){
      cor.test(x$Value_Protein, x$Value_Imaging, method = "spearman", use = "na.or.complete")
      }),
    cor_spearman_tidy = map(cor_spearman, tidy),
    cor_spearman_p = map(cor_spearman_tidy, function(x) x$p.value) %>% unlist, # extract p value
    cor_spearman_padj = map(cor_spearman_p, function(x) p.adjust(x, n = (nrow(dat))/2, "fdr")) %>% unlist, # adjust p value
    cor_spearman_r = map(cor_spearman_tidy, function(x) x$estimate) %>% unlist, # extract R
    #change sign of hpc volume, csf ab42, and mmse
    cor_spearman_r2 = if_else(Imaging %in% c("hp_vol", "csf_ab42_pg_ml__lumip1", "mmse"), cor_spearman_r*-1, cor_spearman_r)
    )
#make glm analysis  
glm_v1 <- 
  dat_prep %>% 
  mutate(
    glm = map(data, function(x){glm(data = x, formula = Value_Imaging~Value_Protein+Gender+age_at_mri+APOE_num)}),
    glm_tidy = map(glm, tidy),
    glm_e = map(glm_tidy, function(x) x$estimate[2]) %>% unlist,
    glm_p = map(glm_tidy, function(x) x$p.value[2]) %>% unlist
    )

#R for plotting ----
pdf(file.path(dir_plots, "Plot_heatmap_r_cor.pdf"), width = unit(20, "cm"), height = unit(20, "cm"))
heatmap_r_v1 <- 
  cor_v1 %>% 
  dplyr::select(Protein, Imaging, cor_spearman_r2) %>% 
  filter(Imaging %in% c(cogn_cor1, imaging_cor1)) %>% 
  tidyheatmaps::tidy_heatmap(rows = Protein, columns = Imaging, values = cor_spearman_r2, cluster_rows = T, cluster_cols = T, fontsize = 6, cellwidth = 8, cellheight = 2, treeheight_row = 20, treeheight_col = 20, show_selected_row_labels = c("Ab42", "pTau217", "pTau181", "pTau231", "TREM2", "GFAP", "NEFL", "MAPT"))
dev.off()
#Padj for plotting
pdf(file.path(dir_plots, "Plot_heatmap_padj_cor.pdf"), width = unit(20, "cm"), height = unit(20, "cm"))
cor_v1 %>% 
  dplyr::select(Protein, Imaging, cor_spearman_p, cor_spearman_padj, cor_spearman_r2) %>% 
  filter(Imaging %in% c(cogn_cor1, imaging_cor1)) %>% 
  #use not adjusted p value
  mutate(nlogpadj = -log10(cor_spearman_p),
         signif = if_else(cor_spearman_p > 0.05, 0, nlogpadj),
         signif = if_else(signif == Inf, 20 , nlogpadj),
         signif2 = if_else(cor_spearman_p > 0.05, NA, nlogpadj),
         signif2 = if_else(signif2 == Inf, 20 , signif2),
         signif2 = if_else(signif2 > 20, 20 , signif2),
         Protein = factor(Protein, levels = heatmap_r_v1$tree_row$labels[heatmap_r_v1$tree_row$order]),
         Imaging = factor(Imaging, levels = heatmap_r_v1$tree_col$labels[heatmap_r_v1$tree_col$order])
         ) %>% 
  arrange(Protein, Imaging) %>% 
  tidyheatmaps::tidy_heatmap(rows = Protein, columns = Imaging, values = signif2, fontsize = 6, cellwidth = 8, cellheight = 2, treeheight_row = 30, treeheight_col = 30, color_na = "grey90", show_selected_row_labels = c("Ab42", "pTau217", "pTau181", "pTau231", "TREM2", "GFAP", "NEFL", "MAPT"), colors = brewer.pal(10, "YlOrRd"))
dev.off()


#Plot values across Braak stages -----
pdf(file.path(dir_plots, "Plot_heatmap_braak.pdf"), width = unit(20, "cm"), height = unit(20, "cm"))
dat %>% 
  filter(tissue == "csf") %>% 
  dplyr::select(data_f1) %>% 
  unnest %>% 
  group_by(protein) %>% 
  filter(!is.na(nav_adni_suvr_fullcg_neocortex)) %>% 
  dplyr::select(protein, braak_stage, Val_z) %>% 
  filter(!is.na(braak_stage), !is.na(Val_z), !Val_z %in% Inf) %>% 
  group_by(braak_stage, protein) %>% 
  summarise(mean = mean(Val_z, na.rm = T)) %>% 
  tidyheatmaps::tidy_heatmap(rows = protein, columns = braak_stage, values = mean, cluster_rows = T, cluster_cols = F, fontsize = 6, cellwidth = 6, cellheight = 4, treeheight_row = 30, treeheight_col = 30, scale = "row")
dev.off()
```
get the clusters from pheatmap
```{r}
library(factoextra)
library(ggdendro)
heatmap_cluster_mtx <- 
  cor_v1 %>% 
  dplyr::select(Protein, Imaging, cor_spearman_r2) %>% 
  filter(Imaging %in% c(cogn_cor1, imaging_cor1)) %>% 
  pivot_wider(names_from = Imaging, values_from = cor_spearman_r2) %>% 
  column_to_rownames("Protein")
#make elbow plot ####
fviz_nbclust(heatmap_cluster_mtx, FUN = hcut, method = "wss") + 
  labs(title = "Elbow Method for Optimal Clusters") +
  plot_layout(widths = unit(5, "cm"), heights = unit(2.5, "cm"))
ggsave(file.path(dir_plots, "plot_elbow.pdf"), useDingbats = F)
#make dendrogram plot ####
plot(hclust(dist(heatmap_cluster_mtx)), main = "Dendrogram", xlab = "Samples", ylab = "Height")
dend_data <- 
  hclust(dist(heatmap_cluster_mtx)) %>% 
  as.dendrogram() %>% 
  dendro_data() 
dend_data %>% 
  ggplot() +
  geom_segment(data = dend_data$segments, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = dend_data$labels, aes(x = x, y = y, label = label), hjust = 1, angle = 90, size = 2.8) +
  labs(title = "Dendrogram") +
  pl_msw0_legend +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  scale_y_continuous(limits = c(-3, 4)) +
  plot_layout(widths = unit(10, "cm"), heights = unit(5, "cm"))
ggsave(file.path(dir_plots, "plot_dendrogram.pdf"), useDingbats = F)
#analyze individual clusters => generate 6 cluster ####
df_heatmap_cluster <- 
  tibble(
    Genes = names(cutree(heatmap_r_v1$tree_row, k = 6)),
    Cluster = cutree(heatmap_r_v1$tree_row, k = 6)) %>% 
  group_by(Cluster) %>% 
  nest()
#visualize the similarities between the clusters ####
pdf(file = file.path(dir_plots, "plot_dissimilarity.pdf"), useDingbats = F)
heatmap_cluster_mtx %>% 
  rownames_to_column("Protein") %>% 
  left_join((df_heatmap_cluster %>% unnest()), by = c("Protein" = "Genes")) %>% 
  column_to_rownames("Protein") %>% 
  aggregate(. ~ Cluster, data = ., FUN = mean) %>% 
  dplyr::select(-Cluster) %>% 
  dist() %>% 
  pheatmap::pheatmap(., cluster_rows = FALSE, cluster_cols = FALSE, cellwidth = 10, cellheight = 10)
dev.off()

heatmap_cluster_mtx %>% 
  rownames_to_column("Protein") %>% 
  left_join((df_heatmap_cluster %>% unnest()), by = c("Protein" = "Genes")) %>% 
  column_to_rownames("Protein") %>% 
  aggregate(. ~ Cluster, data = ., FUN = mean) %>% 
  dplyr::select(-Cluster) %>% 
  dist() %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  write.xlsx(file.path(dir_stats, "zz_af2c.xlsx"), overwrite = T)

#make go term analysis of different clusters ####
df_heatmap_cluster_go <- 
  df_heatmap_cluster %>% 
  filter(!Cluster %in% c(3,6)) %>% 
  mutate(
    go_bp = map(data, function(x) enrichGO(x$Genes, keyType = "SYMBOL", OrgDb = "org.Hs.eg.db", ont = "BP", pvalueCutoff = 1, pAdjustMethod = "fdr")),
    go_bp_res = map(go_bp, function(x) x@result)
    )
```
perform GO term analysis with upregulated genes
```{r, fig.height=10, fig.width=10}
go_v2 <- 
  cor_v1 %>% 
  filter(cor_spearman_padj < 0.05) %>% 
  dplyr::select(Protein, Imaging, cor_spearman_r) %>% 
  group_by(Imaging) %>% 
  nest() %>% 
  mutate(
    cor_pos = map(data, function(x) x %>% filter(cor_spearman_r > 0) %>% pull(Protein)),
    cor_neg = map(data, function(x) x %>% filter(cor_spearman_r < 0) %>% pull(Protein)),
    ) %>% 
  pivot_longer(cols = c(cor_pos, cor_neg), names_to = "Cor", values_to = "Protein") %>% 
  mutate(
    cor_n = map(Protein, length) %>% unlist
    ) %>% 
  filter(cor_n > 5) %>% 
  mutate(
    go_bp = map(Protein, function(x) enrichGO(x, keyType = "SYMBOL", ont = "BP", pvalueCutoff = 1, pAdjustMethod = "fdr", OrgDb = "org.Hs.eg.db")),
    go_cc = map(Protein, function(x) enrichGO(x, keyType = "SYMBOL", ont = "CC", pvalueCutoff = 1, pAdjustMethod = "fdr", OrgDb = "org.Hs.eg.db")),
    go_bp_res = pmap(list(go_bp, Imaging, Cor), function(x,y,z) x@result %>% add_column(var = y, dir = z, go = "bp")),
    go_cc_res = pmap(list(go_cc, Imaging, Cor), function(x,y,z) x@result %>% add_column(var = y, dir = z, go = "cc"))
    )
#check top hits
go_v2 %>% 
  dplyr::select(go_bp_res) %>% 
  unnest() %>% 
  group_by(Imaging, dir) %>% 
  arrange(p.adjust) %>% 
  dplyr::slice(1:20) %>% 
  dplyr::select(dir, everything())
#included go terms
include_GO <- c("neuron apoptotic process", "axonal transport", "modulation of chemical synaptic transmission", "neuroinflammatory response", "glial cell activation")
include_cors <- c("mk_adni_suvr_infcg_braak1", "mk_adni_suvr_infcg_braak2", "mk_adni_suvr_infcg_braak3", "mk_adni_suvr_infcg_braak4", "mk_adni_suvr_infcg_braak5", "mk_adni_suvr_infcg_braak6", "nav_adni_suvr_fullcg_neocortex", "hp_vol", "cdr", "cdr_sob", "mmse")
#make plots
go_v2_pl1 <- 
  go_v2 %>% 
  dplyr::select(go_bp_res) %>% 
  unnest() %>% 
  filter(Description %in% include_GO) %>% 
  dplyr::select(dir, everything()) %>% 
  filter(Imaging %in% include_cors) %>% 
  dplyr::select(dir, Imaging, Description, p.adjust) %>% 
  mutate(
    Imaging = factor(Imaging, levels = include_cors),
    sign = if_else(dir == "cor_pos", 1, -1),
    logpadj = -log10(p.adjust),
    slogpadj = sign(sign) * -log10(p.adjust)
    ) %>% 
  group_by(Description) %>% 
  nest() %>% 
  mutate(
    plot_v1 = pmap(list(data, Description), function(x,y){
      x %>% 
        right_join(tibble(Imaging = include_cors), by = "Imaging") %>% 
        mutate(logpadj = if_else(is.na(logpadj), 0, logpadj),
               slogpadj = if_else(is.na(slogpadj), 0, slogpadj),
               Imaging = factor(Imaging, levels = include_cors)) %>% 
        ggplot(aes(x = Imaging, y = slogpadj, fill = logpadj)) +
        geom_bar(stat = "identity", color = "black", width = 0.7) +
        geom_hline(yintercept = 0, color = "black") +
        geom_hline(yintercept = -log10(0.001), color = "black", linetype = "dashed") +
        geom_hline(yintercept = log10(0.001), color = "black", linetype = "dashed") +
        scale_fill_gradient2(low = "grey80", high = muted("red")) +
        labs(title = y, y = "Signed -log10 Padj") +
        pl_msw0_legend +
        theme(axis.title.x = element_blank()) +
        plot_layout(widths = unit(6, "cm"), heights = unit(2.5, "cm"))
    }),
    dat_exp = pmap(list(data, Description), function(x,y){
      x %>% 
        right_join(tibble(Imaging = include_cors), by = "Imaging") %>% 
        mutate(logpadj = if_else(is.na(logpadj), 0, logpadj),
               slogpadj = if_else(is.na(slogpadj), 0, slogpadj),
               Imaging = factor(Imaging, levels = include_cors))
    })
  )
```
generate score for mitochondrial dysfunction, inflammation, and neuronal cell death, synapse transmission
```{r}
dat_include <- 
  dat %>% 
  filter(tissue == "csf") %>% 
  dplyr::select(data_f1) %>% 
  unnest %>% 
  filter(ATN %in% c("A-T-N-", "A+T-N-", "A+T+N-", "A+T+N+")) 
genelists <- 
  go_v2 %>% 
  dplyr::select(go_bp_res) %>% 
  unnest() %>% 
  ungroup() %>% 
  filter(Description %in% c("neuron apoptotic process", "axonal transport", "modulation of chemical synaptic transmission", "neuroinflammatory response", "glial cell activation")) %>% 
  group_by(Description) %>% 
  nest() %>% 
  mutate(
    genelist = map(data, function(x) x %>% pull(geneID) %>% unlist() %>% paste(collapse = "/") %>% str_split("/") %>% .[[1]] %>% unique()),
    scores = map(genelist, function(x) dat_include %>% filter(protein %in% x) %>% group_by(id) %>% summarise(mean = mean(Val_z, na.rm = T)))
    ) %>% 
  dplyr::select(-data)
dat_gos <- 
  genelists %>% 
  dplyr::select(Description, scores) %>% 
  unnest() %>% 
  left_join(dat_include, by = "id") %>% 
  ungroup() %>% 
  group_by(id, Description) %>% 
  filter(!duplicated(id)) %>% 
  group_by(Description) %>% 
  nest() %>% 
  mutate(
    plot_atn = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = ATN, y = mean, color = ATN)) +
        geom_jitter(width = 0.2, shape = 21) +
        stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, geom = "crossbar", width = 0.3, color = "black", size = 0.3) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm")) +
        stat_compare_means(method = "t.test", hide.ns = T, label = "p.signif", comparisons = FUN.make_comps(dat_include$ATN))
      }),
    stat_atn = map(data, function(x) x %>% filter(mean > -5) %>% compare_means(data = ., formula = mean~ATN, method = "t.test")),
    plot_diagnosis = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = diagnosis_v1, y = mean, color = diagnosis_v1)) +
        geom_jitter(width = 0.2, shape = 21) +
        stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, geom = "crossbar", width = 0.3, color = "black", size = 0.3) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(5:9)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm")) +
        stat_compare_means(method = "t.test", hide.ns = T, label = "p.signif", comparisons = FUN.make_comps(dat_include$diagnosis_v1))
      }),
    stat_dia = map(data, function(x) x %>% filter(mean > -5) %>% compare_means(data = ., formula = mean~diagnosis_v1, method = "t.test")),
    plot_cor_nav = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = nav_adni_suvr_fullcg_neocortex, y = mean)) +
        #geom_jitter(width = 0.2, shape = 21, aes(color = ATN, fill = ATN), alpha = 0.2) +
        geom_smooth(method = "lm", aes(color = ATN, fill = ATN)) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        scale_fill_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    plot_cor_hpc = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = hp_vol, y = mean)) +
        #geom_jitter(width = 0.2, shape = 21, aes(color = ATN, fill = ATN), alpha = 0.2) +
        geom_smooth(method = "lm", aes(color = ATN, fill = ATN)) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        scale_fill_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    plot_cor_braak1 = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = mk_adni_suvr_infcg_braak1, y = mean)) +
       # geom_jitter(width = 0.2, shape = 21, aes(color = ATN, fill = ATN), alpha = 0.2) +
        geom_smooth(method = "lm", aes(color = ATN, fill = ATN)) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        scale_fill_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    plot_cor_braak2 = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = mk_adni_suvr_infcg_braak2, y = mean)) +
        #geom_jitter(width = 0.2, shape = 21, aes(color = ATN, fill = ATN), alpha = 0.2) +
        geom_smooth(method = "lm", aes(color = ATN, fill = ATN)) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        scale_fill_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    plot_cor_braak3 = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = mk_adni_suvr_infcg_braak3, y = mean)) +
        #geom_jitter(width = 0.2, shape = 21, aes(color = ATN, fill = ATN), alpha = 0.2) +
        geom_smooth(method = "lm", aes(color = ATN, fill = ATN)) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        scale_fill_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    plot_cor_braak4 = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = mk_adni_suvr_infcg_braak4, y = mean)) +
        #geom_jitter(width = 0.2, shape = 21, aes(color = ATN, fill = ATN), alpha = 0.2) +
        geom_smooth(method = "lm", aes(color = ATN, fill = ATN)) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        scale_fill_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    plot_cor_braak5 = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = mk_adni_suvr_infcg_braak5, y = mean)) +
        #geom_jitter(width = 0.2, shape = 21, aes(color = ATN, fill = ATN), alpha = 0.2) +
        geom_smooth(method = "lm", aes(color = ATN, fill = ATN)) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        scale_fill_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    plot_cor_braak6 = pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = mk_adni_suvr_infcg_braak6, y = mean)) +
        #geom_jitter(width = 0.2, shape = 21, aes(color = ATN, fill = ATN), alpha = 0.2) +
        geom_smooth(method = "lm", aes(color = ATN, fill = ATN)) +
        pl_msw0_legend_straight +
        scale_color_manual(values = cbPalette[c(1:4)]) +
        scale_fill_manual(values = cbPalette[c(1:4)]) +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    plot_braakstage =pmap(list(data, Description), function(x,y){
      x %>% 
        filter(mean > -5) %>% 
        ggplot(aes(x = braak_stage, y = mean)) +
        geom_point(shape = 21, alpha = 0.5, fill = muted("red")) +
        geom_smooth(color = muted("red"), fill = muted("red")) +
        pl_msw0_legend_straight +
        labs(y = y) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      })
    )
```
make statistical analysis of correlations
```{r, fig.width=10}
#for all participants
cor_go_test <- 
  dat_gos %>% 
  dplyr::select(Description, data) %>% 
  ungroup() %>% 
  unnest() %>% 
  type_convert() %>% 
  pivot_longer(cols = c("mk_adni_suvr_infcg_braak1", "mk_adni_suvr_infcg_braak2", "mk_adni_suvr_infcg_braak3", "mk_adni_suvr_infcg_braak4", "mk_adni_suvr_infcg_braak5", "mk_adni_suvr_infcg_braak6", "nav_adni_suvr_fullcg_neocortex", "hp_vol", "cdr", "cdr_sob", "mmse"), names_to = "params", values_to = "values") %>% 
  group_by(Description, params, ATN) %>% 
  nest() %>% 
  mutate(
    cor.test = map(data, function(x) cor.test(x$mean, x$values, method = "spearman")),
    cor.test.tidy = map(cor.test, tidy),
    cor.r = map(cor.test.tidy, function(x) x$estimate) %>% unlist(),
    cor.p = map(cor.test.tidy, function(x) x$p.value) %>% unlist(),
    cor.padj = map(cor.p, function(x) p.adjust(x, "fdr", n = 11)) %>% unlist()
    )
##save data
cor_go_test %>% 
  dplyr::select(-data, -cor.test, -cor.test.tidy) %>% 
  mutate(across(where(is.numeric), ~ round(x = .x, digits = 3))) %>% 
  write.xlsx(file.path(dir_stats, "stats_go_cor.xlsx"), overwrite = T)

# for > 50 years old participants
cor_go_test_age <- 
  dat_gos %>% 
  dplyr::select(Description, data) %>% 
  ungroup() %>% 
  unnest() %>% 
  type_convert() %>% 
  filter(age_at_mri > 50) %>% 
  pivot_longer(cols = c("mk_adni_suvr_infcg_braak1", "mk_adni_suvr_infcg_braak2", "mk_adni_suvr_infcg_braak3", "mk_adni_suvr_infcg_braak4", "mk_adni_suvr_infcg_braak5", "mk_adni_suvr_infcg_braak6", "nav_adni_suvr_fullcg_neocortex", "hp_vol", "cdr", "cdr_sob", "mmse"), names_to = "params", values_to = "values") %>% 
  group_by(Description, params, ATN) %>% 
  nest() %>% 
  mutate(
    cor.test = map(data, function(x) cor.test(x$mean, x$values, method = "spearman")),
    cor.test.tidy = map(cor.test, tidy),
    cor.r = map(cor.test.tidy, function(x) x$estimate) %>% unlist(),
    cor.p = map(cor.test.tidy, function(x) x$p.value) %>% unlist(),
    cor.padj = map(cor.p, function(x) p.adjust(x, "fdr", n = 11)) %>% unlist()
    )

#make heatmaps ----
pdf(file.path(dir_plots, "heatmap_cor_go.pdf"), useDingbats = F, width = unit(20, "cm"))
cor_go_test %>% 
  dplyr::select(-data, -cor.test, -cor.test.tidy) %>% 
  mutate(params = factor(params, levels = include_cors)) %>% 
  arrange(desc(ATN), params) %>% 
  mutate(r2 = if_else(cor.p < 0.05, cor.r, NA)) %>% 
  tidy_heatmap(rows = Description, columns = c(params, ATN), values = r2, color_na = "white", border_color = "grey80", fontsize = 8, cellwidth = 8, cellheight = 8)
dev.off()
#make heatmaps > 50 ----
pdf(file.path(dir_plots, "heatmap_cor_go_age.pdf"), useDingbats = F, width = unit(20, "cm"))
cor_go_test_age %>% 
  dplyr::select(-data, -cor.test, -cor.test.tidy) %>% 
  mutate(params = factor(params, levels = include_cors)) %>% 
  arrange(desc(ATN), params) %>% 
  mutate(r2 = if_else(cor.p < 0.05, cor.r, NA)) %>% 
  tidy_heatmap(rows = Description, columns = c(params, ATN), values = r2, color_na = "white", border_color = "grey80", fontsize = 8, cellwidth = 8, cellheight = 8)
dev.off()
```
make mediation analysis
```{r}
library(lavaan)
params <- c("age_at_mri", "mmse", "education", "mk_adni_suvr_infcg_metaroi", include_cors)
dat_med <- 
  genelists %>% 
  dplyr::select(Description, scores) %>% 
  unnest() %>% 
  pivot_wider(id_cols = id, names_from = Description, values_from = mean) %>% 
  set_names(c("id", "synapse", "inflammation", "death", "glia", "mito"))
dat_med_all <- 
  dat_include %>% 
  ungroup() %>% 
  filter(id %in% dat_med$id, protein %in% "pTau217") %>% 
  filter(!duplicated(id)) %>% 
  type_convert() %>% 
  dplyr::select(-tissue, -protein) %>% 
  right_join(dat_med, by = "id") %>% 
  mutate(across(.cols = c(params, "synapse", "inflammation", "death", "glia", "mito", "Value_lod"), FUN.scale_z)) %>% 
  filter(synapse > -5, inflammation > -5, death > -5, glia > -5, mito > -5) %>% 
  mutate(hp_vol = hp_vol * -1,
         mmse = mmse * -1) #everything worse should go up
#in a-t-n- and a+t-n-
dat_med_neg <- 
  dat_include %>% 
  ungroup() %>% 
  filter(id %in% dat_med$id, protein %in% "pTau217") %>% 
  filter(!duplicated(id)) %>% 
  type_convert() %>% 
  dplyr::select(-tissue, -protein) %>% 
  right_join(dat_med, by = "id") %>% 
  filter(ATN %in% c("A-T-N-", "A+T-N-")) %>% 
  mutate(across(.cols = c(params, "Value_lod", "synapse", "inflammation", "death", "glia", "mito"), FUN.scale_z)) %>% 
  filter(synapse > -5, inflammation > -5, death > -5, glia > -5, mito > -5) %>% 
  mutate(hp_vol = hp_vol * -1,
         mmse = mmse * -1) #everything worse should go up
```
Mediation model 1: tau => death => inflammation in all and neg
```{r}
model1 <- 
  tibble(var = paste("mk_adni_suvr_infcg_braak", seq(1:6), sep = "")) %>% 
  mutate(
    formula = map(var, function(x)
      paste(
      "# direct effect
      inflammation ~ c1*",x, "+ c2*death + d2*education + d3*age_at_mri
      # mediation
      death ~ b1*",x, "+ e2*education + e3*age_at_mri
      # indirect effect (a*b)
      indirect := b1*c2
      # total effect
      total := c1 + (b1*c2)
      "
      )),
    fit_all = map(formula, function(x) sem(x, data = dat_med_all, bootstrap = 1000)),
    fit_all_summary = map(fit_all, function(x) summary(x, fit.measures = T, standardize = T)),
    fit_all_summary_pe = map(fit_all_summary, function(x) x$pe),
    fit_neg = map(formula, function(x) sem(x, data = dat_med_neg, bootstrap = 1000)),
    fit_neg_summary = map(fit_neg, function(x) summary(x, fit.measures = T, standardize = T)),
    fit_neg_summary_pe = map(fit_neg_summary, function(x) x$pe)
    )
#make plots for all
model1 %>% 
  dplyr::select(var, fit_all_summary_pe) %>% 
  unnest() %>% 
  filter(label %in% c("indirect", "total", "c1", "c2", "b1")) %>% 
  mutate(signif = if_else(pvalue < 0.05, "y", "n"),
         var = factor(var, levels = paste("mk_adni_suvr_infcg_braak", seq(1:6), sep = "")),
         label = factor(label, levels = c("total", "indirect", "c1", "c2", "b1"))) %>% 
  ggplot(aes(x = var, y = std.all)) +
  geom_bar(stat = "identity", position = position_dodge2(), aes(fill = label, color = signif), width = 0.9, alpha = 0.9) +
  scale_fill_manual(values = cbPalette[c(1:5)]) +
  scale_color_manual(values = c(muted("blue"), muted("red"))) +
  scale_y_continuous(limits = c(NA,NA)) +
  pl_msw0_legend +
  plot_layout(widths = unit(6, "cm"), heights = unit(2.5, "cm"))
ggsave(file.path(dir_plots, "plot_mediation_tau_death_inflammation_all.pdf"), useDingbats = F)
```
Mediation model 2: tau => glia => inflammation in all and neg
```{r}
model2 <- 
  tibble(var = paste("mk_adni_suvr_infcg_braak", seq(1:6), sep = "")) %>% 
  mutate(
    formula = map(var, function(x)
      paste(
      "# direct effect
      inflammation ~ c1*",x, "+ c2*glia + d2*education + d3*age_at_mri
      # mediation
      glia ~ b1*",x, "+ e2*education + e3*age_at_mri
      # indirect effect (a*b)
      indirect := b1*c2
      # total effect
      total := c1 + (b1*c2)
      "
      )),
    fit_all = map(formula, function(x) sem(x, data = dat_med_all, bootstrap = 1000)),
    fit_all_summary = map(fit_all, function(x) summary(x, fit.measures = T, standardize = T)),
    fit_all_summary_pe = map(fit_all_summary, function(x) x$pe),
    fit_neg = map(formula, function(x) sem(x, data = dat_med_neg, bootstrap = 1000)),
    fit_neg_summary = map(fit_neg, function(x) summary(x, fit.measures = T, standardize = T)),
    fit_neg_summary_pe = map(fit_neg_summary, function(x) x$pe)
    )
#make plots for all
model2 %>% 
  dplyr::select(var, fit_all_summary_pe) %>% 
  unnest() %>% 
  filter(label %in% c("indirect", "total", "c1", "c2", "b1")) %>% 
  mutate(signif = if_else(pvalue < 0.05, "y", "n"),
         var = factor(var, levels = paste("mk_adni_suvr_infcg_braak", seq(1:6), sep = "")),
         label = factor(label, levels = c("total", "indirect", "c1", "c2", "b1"))) %>% 
  ggplot(aes(x = var, y = std.all)) +
  geom_bar(stat = "identity", position = position_dodge2(), aes(fill = label, color = signif), width = 0.9, alpha = 0.9) +
  scale_fill_manual(values = cbPalette[c(1:5)]) +
  scale_color_manual(values = c(muted("blue"), muted("red"))) +
  scale_y_continuous(limits = c(NA,NA)) +
  pl_msw0_legend +
  plot_layout(widths = unit(6, "cm"), heights = unit(2.5, "cm"))
ggsave(file.path(dir_plots, "plot_mediation_tau_glia_inflammation_all.pdf"), useDingbats = F)
#make plots for neg
model2 %>% 
  dplyr::select(var, fit_neg_summary_pe) %>% 
  unnest() %>% 
  filter(label %in% c("indirect", "total", "c1", "c2", "b1")) %>% 
  mutate(signif = if_else(pvalue < 0.05, "y", "n"),
         var = factor(var, levels = paste("mk_adni_suvr_infcg_braak", seq(1:6), sep = "")),
         label = factor(label, levels = c("total", "indirect", "c1", "c2", "b1"))) %>% 
  ggplot(aes(x = var, y = std.all)) +
  geom_bar(stat = "identity", position = position_dodge2(), aes(fill = label, color = signif), width = 0.9, alpha = 0.9) +
  scale_fill_manual(values = cbPalette[c(1:5)]) +
  scale_color_manual(values = c(muted("blue"), muted("red"))) +
  scale_y_continuous(limits = c(NA,NA)) +
  pl_msw0_legend +
  plot_layout(widths = unit(6, "cm"), heights = unit(2.5, "cm"))
ggsave(file.path(dir_plots, "plot_mediation_tau_glia_inflammation_neg.pdf"), useDingbats = F)
```
Mediation model 3: pTau217 => glia  => inflammation in neg
```{r}
model3 <- 
      "# direct effect
      inflammation ~ c1*Value_lod + c2*glia + d2*education + d3*age_at_mri
      # mediation
      glia ~ b1*Value_lod + e2*education + e3*age_at_mri
      # indirect effect (a*b)
      indirect := b1*c2
      # total effect
      total := c1 + (b1*c2)
      "
fit3 <- sem(model3, data = dat_med_neg, bootstrap = 1000)
summary(fit3, fit.measures = T, standardize = T)
```
Mediation model 4: pTau217 => death  => inflammation in neg
```{r}
model4 <- 
      "# direct effect
      inflammation ~ c1*Value_lod + c2*death + d2*education + d3*age_at_mri
      # mediation
      death ~ b1*Value_lod + e2*education + e3*age_at_mri
      # indirect effect (a*b)
      indirect := b1*c2
      # total effect
      total := c1 + (b1*c2)
      "
fit4 <- sem(model4, data = dat_med_neg, bootstrap = 1000)
summary(fit4, fit.measures = T, standardize = T)
```
Mediation model 5: inflammation => synapse => death in neg
```{r}
model5 <- 
      "# direct effect
      death ~ c1*inflammation + c2*synapse + d2*education + d3*age_at_mri + d4*Value_lod
      # mediation
      synapse ~ b1*inflammation + e2*education + e3*age_at_mri + e4*Value_lod
      # indirect effect (a*b)
      indirect := b1*c2
      # total effect
      total := c1 + (b1*c2)
      "
model5 <- sem(model5, data = dat_med_neg, bootstrap = 1000)
summary(model5, fit.measures = T, standardize = T)
```
Mediation model 6: inflammation => synapse => death in neg
```{r}
model6 <- 
      "# direct effect
      death ~ c1*inflammation + c2*mito + d2*education + d3*age_at_mri + d4*Value_lod
      # mediation
      mito ~b1*inflammation + e2*education + e3*age_at_mri + e4*Value_lod
      # indirect effect (a*b)
      indirect := b1*c2
      # total effect
      total := c1 + (b1*c2)
      "
model6 <- sem(model6, data = dat_med_neg, bootstrap = 1000)
summary(model6, fit.measures = T, standardize = T)
```
Mediation model 7: inflammation => synapse => glia in neg
```{r}
model7 <- 
      "# direct effect
      glia ~ c1*inflammation + c2*synapse + d2*education + d3*age_at_mri +d4*Value_lod
      # mediation
      synapse ~ b1*inflammation + e2*education + e3*age_at_mri + e4*Value_lod
      # indirect effect (a*b)
      indirect := b1*c2
      # total effect
      total := c1 + (b1*c2)
      "
model7 <- sem(model7, data = dat_med_neg, bootstrap = 1000)
summary(model7, fit.measures = T, standardize = T)
```
Mediation model 8: inflammation => transport => glia in neg
```{r}
model8 <- 
      "# direct effect
      glia ~ c1*inflammation + c2*mito + d2*education + d3*age_at_mri +d4*Value_lod
      # mediation
      mito ~ b1*inflammation + e2*education + e3*age_at_mri + e4*Value_lod
      # indirect effect (a*b)
      indirect := b1*c2
      # total effect
      total := c1 + (b1*c2)
      "
model8 <- sem(model8, data = dat_med_neg, bootstrap = 1000)
summary(model8, fit.measures = T, standardize = T)
```
make correlation analysis between ptau 217 and biological themes in neg
```{r}
plot_cor_ptau217 <- 
  dat_med_neg %>% 
  dplyr::select("ATN", "Value_lod", "synapse", "inflammation", "death", "glia", "mito") %>% 
  pivot_longer(cols = synapse:mito) %>% 
  group_by(name) %>% 
  nest() %>% 
  mutate(
    plots = pmap(list(data, name), function(x,y){
      x %>% 
        ggplot(aes(x = Value_lod, y = value)) +
        geom_point(aes(color = ATN), shape = 21) +
        geom_smooth(method = "lm", color = "black") +
        pl_msw0_legend_straight +
        labs(y = y, title = y) +
        scale_color_manual(values = cbPalette[c(1:2)]) +
        stat_cor(size = 2.8) +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      })
    )
```
Look at longitudinal data
```{r}
tmp <- 
  dat_total %>% 
  dplyr::select(id, PID, triad_ID, PID_visit, protein, Value_lod, tissue, diagnosis_v1, plasma_ptau217_pg_ml__alzpath) %>% 
  filter(tissue == "csf") %>% 
  separate(PID_visit, into = c("PID2", "visit"), sep = "VM", remove = F) %>% 
  mutate(visit = if_else(visit == "00", "0", visit)) %>% 
  type_convert() %>% 
  filter(protein == "pTau217") %>% 
  group_by(PID) %>% 
  mutate(n = n())
#define progression by ptau 217 NULISA increase in the csf
tmp_rate <- 
  tmp %>% 
  filter(n > 1, !is.na(visit)) %>% 
  group_by(PID) %>% 
  nest() %>% 
  mutate(
    ptau_baseline = map(data, function(x) x %>% filter(visit %in% min(visit)) %>% pull(Value_lod)) %>% unlist,
    ptau_end = map(data, function(x) x %>% filter(visit %in% max(visit)) %>% pull(Value_lod)) %>% unlist,
    time_baseline = map(data, function(x) x %>% filter(visit %in% min(visit)) %>% pull(visit)) %>% unlist,
    time_end = map(data, function(x) x %>% filter(visit %in% max(visit)) %>% pull(visit)) %>% unlist,
    time_diff = (time_end - time_baseline)/12,
    ptau_diff = ptau_end - ptau_baseline,
    ptau_diff_year = ptau_diff / time_diff
    ) %>% 
  filter(time_diff != 0) %>% 
  dplyr::select(-data)
#generate a merged dataframe
dat_long_merge <- 
  dat_gos %>% 
  dplyr::select(Description, data) %>% 
  unnest() %>% 
  dplyr::select(Description, id, PID, mean, diagnosis_v1, ATN, age_at_mri) %>% 
  left_join(tmp_rate, by = "PID") %>% 
  filter(!is.na(ptau_diff)) %>% 
  group_by(Description) %>% 
  mutate(n = n())

#make dataframe of all samples that were measured ----
dat_all_tmp <- 
  dat %>% 
  filter(tissue == "csf") %>% 
  dplyr::select(data) %>% 
  unnest %>% 
  dplyr::select(id, PID, triad_ID, PID_visit, protein, Val_z, Value_lod, tissue, diagnosis_v1, ATN, Gender, age_at_mri) %>% 
  separate(PID_visit, into = c("PID2", "visit"), sep = "VM", remove = F) %>% 
  mutate(visit = if_else(visit == "00", "0", visit)) %>% 
  type_convert()
#make score for all individuals ----
dat_all_scores <- 
  genelists %>% 
  mutate(
    scores = map(genelist, function(x) dat_all_tmp %>% filter(protein %in% x) %>% group_by(id) %>% summarise(mean = mean(Val_z, na.rm = T)))
    ) %>% 
  dplyr::select(-genelist) %>% 
  unnest() %>% 
  pivot_wider(names_from = Description, id_cols = id, values_from = mean)

#longitudinal data for CU ----
dat_prep_long_CU <- 
  dat_all_tmp %>% 
  ungroup() %>% 
  filter(protein == "pTau217") %>% 
  group_by(PID) %>% 
  mutate(n = n()) %>% 
  filter(n > 1) %>% 
  filter(!diagnosis_v1 %in% c("MCI", "AD")) %>% 
  left_join(dat_all_scores, by = "id") %>% 
  group_by(PID) %>% 
  mutate(glia_diff = `glial cell activation` - `glial cell activation`[visit == min(visit)],
         synapse_diff = `modulation of chemical synaptic transmission` - `modulation of chemical synaptic transmission`[visit == min(visit)],
         infl_diff = `neuroinflammatory response` - `neuroinflammatory response`[visit == min(visit)],
         death_diff = `neuron apoptotic process` - `neuron apoptotic process`[visit == min(visit)],
         transp_diff = `axonal transport` - `axonal transport`[visit == min(visit)],
         ptau_diff = Value_lod - Value_lod[visit == min(visit)],
         visit_y = visit / 12
         ) %>% 
  pivot_longer(cols = glia_diff:ptau_diff, names_to = "Params", values_to = "Value") %>% 
  group_by(Params) %>% 
  nest() %>% 
  mutate(
    data = map(data, function(x){
      x %>% 
        group_by(triad_ID) %>% 
        mutate(baseline = Val_z[visit == min(visit)]) %>% 
        ungroup()
      }),
    plots = pmap(list(data, Params), function(x,y){
      x %>% 
        ggplot(aes(x = visit_y, y = Value, color = ATN3)) +
        geom_smooth(method = "lm", se = T, alpha = 0.3) +
        scale_color_manual(values = cbPalette[c(1,2)]) +
        labs(y = y, title = y) +
        pl_msw0_legend_straight +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    stats = map(data, function(x) lmer(Value ~ visit_y*ATN3 + Gender + age_at_mri + baseline + (1 | PID), data = x)),
    stats_summary = map(stats, summary),
    stats_tidy = map(stats, tidy),
    #check only for A-
    stats_Aneg = map(data, function(x) x %>% filter(ATN3 == "A-") %>% lmer(Value ~ visit_y + Gender + age_at_mri + baseline + (1 | PID), data = .)),
    stats_Aneg_tidy = map(stats_Aneg, tidy),
    #check only for A+
    stats_Apos = map(data, function(x) x %>% filter(ATN3 == "A+") %>% lmer(Value ~ visit_y + Gender + age_at_mri + baseline + (1 | PID), data = .)),
    stats_Apos_tidy = map(stats_Apos, tidy),
    )
#longitudinal data for CI ----
dat_prep_long_CI <- 
  dat_all_tmp %>% 
  ungroup() %>% 
  filter(protein == "pTau217") %>% 
  group_by(PID) %>% 
  mutate(n = n()) %>% 
  filter(n > 1) %>% 
  filter(diagnosis_v1 %in% c("MCI", "AD")) %>% 
  left_join(dat_all_scores, by = "id") %>% 
  group_by(PID) %>% 
  mutate(glia_diff = `glial cell activation` - `glial cell activation`[visit == min(visit)],
         synapse_diff = `modulation of chemical synaptic transmission` - `modulation of chemical synaptic transmission`[visit == min(visit)],
         infl_diff = `neuroinflammatory response` - `neuroinflammatory response`[visit == min(visit)],
         death_diff = `neuron apoptotic process` - `neuron apoptotic process`[visit == min(visit)],
         transp_diff = `axonal transport` - `axonal transport`[visit == min(visit)],
         ptau_diff = Value_lod - Value_lod[visit == min(visit)],
         visit_y = visit / 12
         ) %>% 
  pivot_longer(cols = glia_diff:ptau_diff, names_to = "Params", values_to = "Value") %>% 
  group_by(Params) %>% 
  nest() %>% 
  mutate(
    data = map(data, function(x){
      x %>% 
        group_by(triad_ID) %>% 
        mutate(baseline = Val_z[visit == min(visit)]) %>% 
        ungroup()
      }),
    plots = pmap(list(data, Params), function(x,y){
      x %>% 
        ggplot(aes(x = visit_y, y = Value, color = ATN3)) +
        geom_smooth(method = "lm", se = T, alpha = 0.3) +
        scale_color_manual(values = cbPalette[c(1,2)]) +
        labs(y = y, title = y) +
        pl_msw0_legend_straight +
        plot_layout(widths = unit(2.5, "cm"), heights = unit(2.5, "cm"))
      }),
    stats = map(data, function(x) lmer(Value ~ visit_y*ATN3 + Gender + age_at_mri + baseline + (1 | PID), data = x)),
    stats_summary = map(stats, summary),
    stats_tidy = map(stats, tidy),
    stats_Apos = map(data, function(x) x %>% filter(ATN3 == "A+") %>% lmer(Value ~ visit_y + Gender + age_at_mri + baseline + (1 | PID), data = .)),
    stats_Apos_tidy = map(stats_Apos, tidy),
    )
```